Untwisting the Tetrahelix (v0.4)

### 2017-02-16 doc/TetrahelixUntwist.tex
major simplfication, and some complexity

The Boerdijk--Coxeter helix\cite{coxeter1985simplicial} (BC helix),
is a face-to-face stack of tetrahedra that winds about a straight axis.
The vertices of the tetrahedra
lie upon three
helices about the central axis.
The Tetrobot/Glussbot\cite{TetrobotBook} project
uses the regularity of this geometry to make a tentacle-like robot that can crawl like a slug or mollusc.
The Tetrobot concept
is to use mechanical members, called actuators, which can change their length, connected by special joints, called the Song-Kwon-Kim\cite{song2003spherical} or turret joint,
which allow many
members to come to a single point.
Such machines can follow purely regual mathematical models such as the Boerdijk–-Coxeter helix or the Octet Truss\cite{richard1961synergetic}.
Because architects, structural engineers, and robotocists are inspired by and follow such mathematical models but can build
structures and machines of differing or even dynamically changing length, it is useful to develop
the mathematics of structure formed from tetrahedra where we relax regularity. Buckminster Fuller called the BC helix a \emph{tetrahelix}\cite{fuller1982synergetics},
a term now commonly used. In this paper we reserve BC helix to mean the purely regular structre and use \emph{tetrahelix} to refer
to any structure isomophic to a the BC helix, whether regular or not.

@@

\[
H_{BCsymmmetric}(n,c) =
\left [
  \begin{tabular}{c}
   $ r  \cos(\rho_{bc} \kappa  + c 2 \pi /3) $\\
   $ r  \sin(\rho_{bc} \kappa  + c 2 \pi /3) $\\
   $ \kappa 3  d $
  \end{tabular}
  \right ],
\text{where:}
  \begin{tabular}{c}
 $\kappa = n + c/3$ \\
    $\rho_{bc} = (3 \theta - 2 \pi)$ \\
   $ \theta = \arccos(-2/3) $
  \end{tabular}      
\]

@@

Checking the important special case of the BC helix, we find that this equation
indeed holds true (treating $d$ in this equation as $3 h_{bc}$ as defined by
Gray and Coxeter), where they are using it for the axial height from one node to
the next of a different color, but we use it to mean distance for the same color.

@@

In choosing $\rho$, one greatly constrains $r$ and $h$, but does not completely
determine both of them together. In the formulation below we assume that
the choice of $d_{\rho}$ is more convenient than choosing the radius,
but that is somewhat arbitrary.

Rewriting our formulation in terms of $\rho$:
\[
H_{general}(n,c,\rho,d_{\rho},r_{\rho}) =
\left [
  \begin{tabular}{c}
   $ r_{\rho} \cos(\frac{\rho \kappa}{2\pi} + c 2 \pi /3) $\\
   $ r_{\rho}  \sin(\frac{\rho \kappa}{2\pi} + c 2 \pi /3) $\\
   $ d_{\rho} \kappa $
  \end{tabular}
  \right ],
\text{where:}
\begin{tabular}{c}
  $   1 = d_{\rho}^2 + 4 r_{\rho}^2 (\sin{ \rho / 2})^2 $ \\
    $\kappa = n + c/3$ \\
  \end{tabular}      
\]

@@

\begin{theorem}
  \label{generalformulaoptimal}
  The tetrahelices generated by $H_{general}$ are optimal in terms of minimum maximum member length, minimum total member length, and minimum
  sum of squared lengths.
\end{theorem}

\begin{proof}
  This requires poof.
\end{proof}

By Theoerm \ref{eventhirds}, we can compute the (at most) three edge-lengths of an optimal
tetrahelix by (where $dist$ is the cartesian distance function):
\begin{align*}
  \text{rail} &= dist(H_{general}(n,c,\rho,h_{\rho}),H_{general}(n+1,c,\rho,d_{\rho})) = 1 \\
  \text{one-hop} &= dist(H_{general}(n,c,\rho,h_{\rho}),H_{general}(n,c+1,\rho,d_{\rho}))  \\
  \text{two-hops} &= dist(H_{general}(n,c,\rho,h_{\rho}),H_{general}(n,c+2,\rho,d_{\rho}))  \\  
\end{align*}
Which are invarinat for all $n$ and $c$.

\begin{align*}
  \text{one-hop} &= dist(H_{general}(n,c,\rho,h_{\rho}),H_{general}(n,c+1,\rho,d_{\rho}))  \\
  \text{one-hop} &= dist(H_{general}(0,0,\rho,h_{\rho}),H_{general}(0,1,\rho,d_{\rho}))  \\  
  \text{one-hop}  &= \sqrt{d_{\rho}^2/9 + (r\sin{\frac{\rho(3\cdot 0 + 0)}{2\pi}} - r\sin{\frac{\rho(3\cdot 0 + 0+1)}{2\pi}})^2  +
    (r\cos{\frac{\rho(3\cdot0+0)}{2\pi}} - r\cos{\frac{\rho(3\cdot + 0+1)}{2\pi}})^2
  \text{one-hop}  &= \sqrt{d_{\rho}^2/9 + (r\sin{\frac{\rho(0)}{2\pi}} - r\sin{\frac{\rho(1)}{2\pi}})^2  +
    (r\cos{\frac{\rho(0)}{2\pi}} - r\cos{\frac{\rho(1)}{2\pi}})^2
  } \\  
\end{align*}

@@

However, choosing $d = 1$ and $\rho = 0$ does fix the radius. We would like to choose the radius so that the
minimax difference between the lengths is as small as possible.

@@

When $\rho = 0$, our formulae for distances are particular simple. Let $a$ be the length of the equilateral triangle
forming the beam cross section. Then by geometry $r = a \sqrt{3}$.
\begin{align*}
  \text{rail} &=  1 \\
  \text{one-hop} &= \sqrt{\frac{1}{3}^2 + (\frac{3a}{2\sqrt{3}})^2 + (a/2)^2}\\
\text{one-hop}  &= \sqrt{\frac{1}{9} + a^2} \\
    \text{two-hops} &= \sqrt{\frac{2}{3}^2 + (\frac{3a}{2\sqrt{3}})^2 + (a/2)^2}  \\
\text{two-hops}    &= \sqrt{\frac{4}{9} + a^2}
\end{align*}
Computing the derivative of two-hops - one-hop:
\begin{align*}
 \frac{\partial \text{two-hops} - \text{one-hop}}{a} &= \frac{\partial \sqrt{4/9 + a^2} - \sqrt{1/9 + a^2}}{\partial a} \\
  \frac{\partial \text{two-hops} - \text{one-hop}}{a} &= \frac{a}{\sqrt{a^2 + 4/9}} - \frac{a}{\sqrt{ a^2 + 1/9}} 
\end{align*}
which is negative for all $a> 0$. Thererfore two-hop - one-hop
is minimized when we maximize $a$, up until one-hop = rail.

Note that two-hops increases more and more slowly as $a$ increases. Therefore the minimum maximum will
be obtained with the value of $a$ that when $one-hop = 1$. This can be observed from the
plot of two-hops - one-hops as a function of $a$, or by observing that the derivate of two-hops - one-hop
is negative for $a > 0$. (That is, the value of one-hop increases faster than two-hops.)

We seek the value of $a$ that minimizes the maximum difference between these three lengths. We find
by numerical experimentation that this difference is minimized that the one-hop distance is equal to $1$,
which also has the pleasing property of giving us only two lengths. Then
\begin{align*}
   1  &=  \sqrt{\frac{1}{9} + a^2} \\
     a  &= \sqrt{\frac{8}{9}} 
\end{align*}

Therefore $ r = \sqrt{\frac{8}{27}} \approx 0.5443 $, producing length difference in the two-hop rail of
$\frac{2}{\sqrt{3}}$
and $1$ of $\approx 15.47\% $.

Note: Another interesting solution is derived by setting (one-hop + two-hop)/2 = 1,  occurs at $a = \sqrt{35/3}/4$,
which produces three length classes of $11/12, 12/12, 13/12$.

\begin{conjecture}
  The equitetrabeam is also optimal under the metric of the total length of all members 
  for an equilateral tetrabeam isomorphic to the Boerdijk--Coxeter helix.
\end{conjecture}

\end{proof}

@@

Note: Need to understand how pitch is computed from negative lambda values.

@@

@@

>>>>

The Boerdijk--Coxeter helix\cite{coxeter1985simplicial} (BC helix),
is a face-to-face stack of tetrahedra that winds about a straight axis.
Because architects, structural engineers, and robotocists are inspired by and follow such mathematical models but can build
structures and machines of differing or even dynamically changing length, it is useful to develop
the mathematics of structure formed from tetrahedra where we relax regularity.
The vertices of the tetrahedra
lie upon three
helices about the central axis.
The Tetrobot/Glussbot\cite{TetrobotBook} project
uses the regularity of this geometry to make a tentacle-like robot that can crawl like a slug or mollusc.
The Tetrobot concept
is to use mechanical members, called actuators, which can change their length, connected by special joints, called the Song-Kwon-Kim\cite{song2003spherical} or turret joint,
which allow many
members to come to a single point.
Such machines can follow purely regual mathematical models such as the Boerdijk–-Coxeter helix or the Octet Truss\cite{richard1961synergetic}.

Buckminster Fuller called the BC helix a \emph{tetrahelix}\cite{fuller1982synergetics},
a term now commonly used. In this paper we reserve BC helix to mean the purely regular structre and use \emph{tetrahelix} to refer
to any structure isomophic to a the BC helix, whether regular or not.

@@

\[
H_{BCsymmmetric}(n,c) =
\left [
  \begin{tabular}{c}
   $ r  \cos(\rho_{bc} \kappa  + c 2 \pi /3) $\\
   $ r  \sin(\rho_{bc} \kappa  + c 2 \pi /3) $\\
   $ \kappa 3  h_{bc} $
  \end{tabular}
  \right ],
\text{where:}
  \begin{tabular}{c}
 $\kappa = n + c/3$ \\
    $\rho_{bc} = (3 \theta - 2 \pi)$ \\
    $ \theta = \arccos(-2/3) $
    $ h_{bc} = 1/\sqrt{10} $ \\    
  \end{tabular}      
\]

@@

Checking the important special case of the BC helix, we find that this equation
indeed holds true (treating $d$ in this equation as $3 h_{bc}$ as defined by
Gray and Coxeter, that is, $d_{bc} = 3h_{bc}$, where they are using it for the axial height from one node to
the next of a different color, but we use it to mean distance for the same color.

@@

In choosing $\rho$, one greatly constrains $r$ and $h$, but does not completely
determine both of them together, so we treat both as parameters.

Rewriting our formulation in terms of $\rho$:
\[
H_{general}(n,c,\rho,d_{\rho},r_{\rho}) =
\left [
  \begin{tabular}{c}
   $ r_{\rho} \cos(\rho \kappa + c 2 \pi /3) $\\
   $ r_{\rho}  \sin(\rho \kappa + c 2 \pi /3) $\\
   $ d_{\rho} \kappa $
  \end{tabular}
  \right ],
\text{where:}
\begin{tabular}{c}
  $   1 = d_{\rho}^2 + 4 r_{\rho}^2 (\sin{ \rho / 2})^2 $ \\
    $\kappa = n + c/3$ \\
  \end{tabular}      
\]

@@

\begin{theorem}
  \label{generalformulaoptimal}
  The tetrahelices generated by $H_{general}$ are optimal in terms of minimum maximum member length when $r_{\rho}$ is chosen so that
  the length of the one-hop edge is equal to the rail length.
\end{theorem}

\begin{proof}
  This requires a minimax argument.
\end{proof}

By Theoerm \ref{eventhirds}, we can compute the (at most) three edge-lengths of an optimal
tetrahelix by (where $dist$ is the cartesian distance function):
\begin{align*}
  \text{rail} &= dist(H_{general}(n,c,\rho,d_{\rho},r_{\rho}),H_{general}(n+1,c,\rho,d_{\rho}),r_{\rho})) = 1 \\
  \text{one-hop} &= dist(H_{general}(n,c,\rho,d_{\rho},r_{\rho}),H_{general}(n,c+1,\rho,d_{\rho},r_{\rho}))  \\
  \text{two-hops} &= dist(H_{general}(n,c,\rho,d_{\rho},r_{\rho}),H_{general}(n,c+2,\rho,d_{\rho},r_{\rho}))  \\  
\end{align*}
Which are invarinat for all $n$ and $c$.

\begin{align*}
  \text{one-hop} &= dist(H_{general}(n,c,\rho,h_{\rho}),H_{general}(n,c+1,\rho,d_{\rho}),r_{\rho})  \\
  \text{one-hop} &= dist(H_{general}(0,0,\rho,h_{\rho}),H_{general}(0,1,\rho,d_{\rho},r_{\rho}))  \\  
  \text{one-hop}  &= \sqrt{\frac{d_{\rho}^2}{9} + (,r_{\rho}\sin(\rho(3\cdot 0 + 0)) - r_{\rho}\sin(\rho(3\cdot 0 + 1)+\frac{2\pi}{3}))^2  +
    (r_{\rho}\cos(\rho(3\cdot0+0)) - r_{\rho}\cos(\rho(3\cdot 0+1) + \frac{2\pi}{3})))^2} \\
  \text{one-hop}  &= \sqrt{\frac{d_{\rho}^2}{9} + (0  - r_{\rho}\sin(\rho + \frac{2\pi}{3}))^2  +
    (r_{\rho} - r_{\rho}\cos(\rho + \frac{2\pi}{3}))^2} \\
  \text{one-hop}  &= \sqrt{\frac{d_{\rho}^2}{9} + r_{\rho}^2\sin^2(\rho + \frac{2\pi}{3})  +
    r_{\rho}^2(1 - \cos(\rho + \frac{2\pi}{3}))^2} \\
  \text{one-hop}  &= \sqrt{\frac{d_{\rho}^2}{9} + r_{\rho}^2(\sin^2(\rho + \frac{2\pi}{3})  + (1 - \cos(\rho + \frac{2\pi}{3}))^2)} \\
\end{align*}

By similar algebra and trigonometry:
\begin{align*}
  \text{two-hops}  &= \sqrt{\frac{4d_{\rho}^2}{9} + r_{\rho}^2(\sin^2(\rho + \frac{2\pi}{3})  + (1 - \cos(\rho + \frac{2\pi}{3}))^2)} \\
\end{align*}

Let $f_{\rho} = \sin^2(\rho + \frac{2\pi}{3})  + (1 - \cos(\rho + \frac{2\pi}{3})^2$.

The expression $f_{\rho}$ is maximized when $\rho = 2\pi/6 = 60\degree$, so that
is in some sense the ``wildest'' rail angle which maximizes the lengths. Since this expression is increasing up until this angle,
we conclude that:
\begin{align*}
  \text{two-hops} - \text{one-hop}  &= \sqrt{\frac{4d_{\rho}^2}{9} + r_{\rho}^2f_{\rho}} - \sqrt{\frac{d_{\rho}^2}{9} + r_{\rho}^2f_{\rho}} 
\end{align*}

The analysis below here is incorrect because it doesn't take into account the change in $d_{\rho}$ with $\rho$. In the special case that
$\rho = 0$, the analysis is correct. Possibly I need to change to use two equations to try to find the optimal $\rho$ here.
In any case it must unify with the BC math!

Tomorrow try combining with our equation for $d_{\rho}$ to see if I can make it come out.

Since $d_{\rho}$ and $r_{\rho}$ are positive, this difference decreases as $f_{\rho}$ increases. Therefore we decrease the minimax length
of the whole system as we increase $\rho$ up to $\pi/3$ up unto the point that the shorter, one-hop distance is equal to the rail-length ($1$).
Therefore, to optimize the whole system so long as $\rho < \pi/3$, we equate one-hop to $1$ to find the optimum radius:

\begin{align*}
  1 &= \sqrt{\frac{d_{\rho}^2}{9} + r_{\rho}^2(\sin^2(\rho + \frac{2\pi}{3})  + (1 - \cos(\rho + \frac{2\pi}{3}))^2)} \\
  r_{\rho} &= \frac{\sqrt{9-d^2}  }{3\sqrt{\sqrt{3}\sin(\rho) +\cos(\rho) + 2}}
\end{align*}
(where $\sqrt{3}\sin(\rho) +\cos{\rho} + 2 \neq 0$), which does not occur when $\rho \leq \pi/3$).

Therefore, by computing $r_{\rho}$ from this equation, we can construct minimax optimal tetrahelix for an $\rho < \pi/3$.

(Note: This fails the check of computing the BC radius for the BC rail angle.)

@@

However, choosing $d = 1$ and $\rho = 0$ does not fix the radius. We would like to choose the radius so that the
minimax difference between the lengths is as small as possible.

@@

When $\rho = 0$, our formulae for distances are particular simple. Let $a$ be the length of the equilateral triangle
forming the beam cross section. Then by geometry $r = a \sqrt{3}$.
\begin{align*}
  \text{rail} &=  1 \\
  \text{one-hop} &= \sqrt{\frac{1}{9} + 3r^2}\\
\text{two-hops}    &= \sqrt{\frac{4}{9} + 3r^2}
\end{align*}
Computing the derivative of two-hops - one-hop:
\begin{align*}
 \frac{\partial \text{two-hops} - \text{one-hop}}{r} &= \frac{\partial \sqrt{4/9 + 3r^2} - \sqrt{1/9 + 3r^2}}{\partial r} \\
  \frac{\partial \text{two-hops} - \text{one-hop}}{r} &= 9r (\frac{1}{\sqrt{27r^2 + 4}} - \frac{1}{\sqrt{ 27r^2 + 1}}) 
\end{align*}
which is negative for all $r> 0$. Thererfore two-hop - one-hop
is minimized when we maximize $r$, up until one-hop = rail.

Note that two-hops increases more and more slowly as $r$ increases. Therefore the minimum maximum will
be obtained with the value of $r$ that when $one-hop = 1$. This can be observed from the
plot of two-hops - one-hops as a function of $r$, or by observing that the derivate of two-hops - one-hop
is negative for $r > 0$. (That is, the value of one-hop increases faster than two-hops.) 
Setting the one-hop distance to $d = 1$ also has the pleasing property of giving us only two lengths. Then
\begin{align*}
   1  &=  \sqrt{\frac{1}{9} + 3r^2} \\
   r  &= \sqrt{\frac{8}{27}} \\
   r &\approx 0.5443
\end{align*}

This radius produces a two-hop rail length of
$\frac{2}{\sqrt{3}}$. The difference between this 
and $1$ of $\approx 15.47\% $.

Note: Another interesting solution is derived by setting (one-hop + two-hop)/2 = 1,  occurs at $r = \sqrt{35}/4$,
which produces three length classes of $11/12, 12/12, 13/12$.

@@

@@

See if we can accurately compute the derivative of two-hop - one-hop and if it is always negative, meaning that
it is always optimal to set one-hop = 1.  The we have a way to compute one-hope.

@@

\begin{align*}
  \text{rail} &=  1 \\
  \text{one-hop} &= \sqrt{\frac{1}{3}^2 + (\frac{3a}{2\sqrt{3}})^2 + (a/2)^2}\\
\text{one-hop}  &= \sqrt{\frac{1}{9} + a^2} \\
    \text{two-hops} &= \sqrt{\frac{2}{3}^2 + (\frac{3a}{2\sqrt{3}})^2 + (a/2)^2}  \\
\text{two-hops}    &= \sqrt{\frac{4}{9} + a^2}
\end{align*}
Computing the derivative of two-hops - one-hop:
\begin{align*}
 \frac{\partial \text{two-hops} - \text{one-hop}}{a} &= \frac{\partial \sqrt{4/9 + a^2} - \sqrt{1/9 + a^2}}{\partial a} \\
  \frac{\partial \text{two-hops} - \text{one-hop}}{a} &= \frac{a}{\sqrt{a^2 + 4/9}} - \frac{a}{\sqrt{ a^2 + 1/9}} 
\end{align*}
